>

### 2019-05-29 paper/spinifex-paper.tex
di tweaks on plane

\begin{itemize}
\tightlist
\item
  \(\textbf{X}\), the data, an \(n \times p\) numeric matrix to be
  projected to 2D.
\item
  \(\textbf{B} = (B_1,~ B_2)\), any 2D orthonormal projection basis,
  \(p \times 2\) matrix, describing the projection from \(p\) to two
  dimensions. We will call this the ``original projection'' because it
  is the starting point for the manual tour.
\item
  \(k\), is the index of the variable to manipulate, called the ``manip
  var''.
\item
  \(|\textbf{B}_k|\), the norm of the manip var, \(k\), the length of
  this basis vector between 0 and 1 inclusive.
\item
  \(\textbf{e}\), a 1D coordinate zero basis vector of length \(p\) with
  the \(k\)-th element set to one.
\item
  \(\textbf{M} = orthonormalize(\textbf{B}|\textbf{e})\), the
  manipulation space, a \(p \times 3\) matrix where out-of-plane
  rotation takes place.
\item
  \(\theta\), the angle of in-projection-plane rotation, for example, on
  the reference axes; \(c_\theta, s_\theta\) are the cosine and sine.
\item
  \(\phi\), the angle of out-of-projection-plane rotation, into the
  manipulation space; \(c_\phi, s_\phi\) are the cosine and sine.
\item
  \(\textbf{U}\), the axis of rotation for \(\phi\).
\item
  \(\textbf{R}\), the rotation matrix, containing the rotaions of
  \(\theta\) and \(\phi\) within the manipulation space, \(\textbf{M}\).
\item
  \(\textbf{RM} = \textbf{M} * \textbf{R}\), the result of the
  manipulation space after the application of the rotation matrix.
\item
  \(\textbf{P} = \textbf{X} * \textbf{RM}\), the data projection, the
  data as projected through the rotated manipulation space.
\end{itemize}

@@

\hypertarget{step-3-generate-rotation}{%
\subsubsection{Step 3) Generate
rotation}\label{step-3-generate-rotation}}

Imagine moving the axis of manip var, the red line in Figure
\ref{fig:step2}, like a lever. This is effectively the process of the
radial manual control, that rotates the manip variable into and out of
the 2D projection (Figure \ref{fig:step3}). As the variable is
controlled, the manipulation space rotates, and the projection onto the
horizontal projection plane correspondingly changes. This is a manual
tour. Generating a sequence of values for the horizontal and vertical
angles produces a path for the rotation of the manipulation space. This
defines the (orthonormally-constrained) rotation on the basis
coefficients.

For a radial tour, fix the (horizontal) angle within the projection
plane, \(\theta\), and define a sequence for the (vertical) angle,
\(\phi\), defining movement out of the plane, that changes the initial
value of \(|\textbf{B}_k|\) to a maximum of 1, and then to 0, before
returning to the initial value.

Rotations are defined by the axes they pivot on, \(\theta\) then is the
rotation around the \(Z\) axis. While \(\phi\) will be the rotation
around, \(\textbf{U}\), a new axis on the \(XY\) plane, orthogonal to
\(\phi\). Given these axes, the SO(3) rotation matrix can be derived by
the following.

@@

\hypertarget{creating-the-animation}{%
\subsection{Creating the animation}\label{creating-the-animation}}

The steps outlined above can be used to create any arbitrary rotation in
the manip space. To use these for sensitivity analysis, the radial
rotation is built into an animation where the manip var is rotated fully
into the projection, completely out, and then back to the initial value.
This involves changing \(\phi\), the angle relative to its initial
value. Apply the transformation \(\phi_i \leftarrow \phi_i-\phi_1\) such
that \(\phi\) is relative to the projection plane. When \(\phi\) is 0,
\(|\textbf{B}_k| = 1\), and when \(\phi\) is \(\pi/2\), the manip var
has no contribution to the projection.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Set initial value to be \(\phi_1\).
\item
  Set an angle value (\(\Delta_\phi\)) that sets the step size to make
  in rotating the manip var into and out of the projection. (Note: Using
  angle value to control this is consistent with the tour algorithm, as
  implemented in the \pkg{tourr}).
\item
  Step towards \(0\), where the manip var has a completely in the
  projection.
\item
  Step towards \(\pi/2\) where the manip variable has no contribution.
\item
  Step back to \(\phi_1\).
\end{enumerate}

@@

\hypertarget{checks-that-need-to-be-made}{%
\subsubsection{Checks that need to be
made}\label{checks-that-need-to-be-made}}

The value of \(\phi\) is relative to its initial value. The
transformation \(\phi_i\) - \(\phi_1\) is useful to think about \(\phi\)
relative to the basis plane. Additionally, the value of \(\phi\) may be
out of phase by a factor of \(\pi\). If the rotation does not behave as
expected these are the first places to check.

>>>>

\begin{itemize}
\tightlist
\item
  \(\textbf{X}\), the data, an \(n \times p\) numeric matrix to be
  projected to 2D.
\item
  \(\textbf{B} = (B_1,~ B_2)\), any 2D orthonormal projection basis,
  \(p \times 2\) matrix, describing the projection from \(p\) to two
  dimensions. We will call this the ``original projection'' because it
  is the starting point for the manual tour.
\item
  \(k\), is the index of the variable to manipulate, called the ``manip
  var''.
\item
  \(|\textbf{B}_k|\), the norm of the manip var, \(k\), the length of
  this basis vector between 0 and 1 inclusive.
\item
  \(\textbf{e}\), a 1D coordinate zero basis vector of length \(p\) with
  the \(k\)-th element set to one.
\item
  \(\textbf{M} = orthonormalize(\textbf{B}|\textbf{e})\), the
  manipulation space, a \(p \times 3\) matrix where out-of-plane
  rotation takes place.
\item
  \(\theta\), the angle of in-projection-plane rotation, for example, on
  the reference axes; \(c_\theta, s_\theta\) are the cosine and sine.
\item
  \(\phi\), the angle of out-of-projection-plane rotation, into the
  manipulation space; \(c_\phi, s_\phi\) are the cosine and sine. The
  initial value for animation purposes is \(\phi_1\).
\item
  \(\textbf{U}\), the axis of rotation for \(\phi\).
\item
  \(\textbf{R}\), the rotation matrix, containing the rotaions of
  \(\theta\) and \(\phi\) within the manipulation space, \(\textbf{M}\).
\item
  \(\textbf{RM} = \textbf{M} * \textbf{R}\), the result of the
  manipulation space after the application of the rotation matrix.
\item
  \(\textbf{P} = \textbf{X} * \textbf{RM}\), the data projection, the
  data as projected through the rotated manipulation space.
\end{itemize}

XXX Clean up the notation, not all of this is needed or useful. The
shadow of \(e\) in the projection may need some notation. The projected
data should be \(Y\) not \(P\).

@@

\hypertarget{step-3-defining-a-3d-rotation}{%
\subsubsection{Step 3) Defining a 3D
rotation}\label{step-3-defining-a-3d-rotation}}

The basis vector corresponding to the manip var (red line in Figure
\ref{fig:step2}), can be operated like a lever. This is the process of
the manual control, that rotates the manip variable into and out of the
2D projection (Figure \ref{fig:step3}). As the variable is controlled,
the manipulation space rotates, and the projection onto the horizontal
projection plane correspondingly changes. This is a manual tour.
Generating a sequence of values for the horizontal and vertical angles
produces a path for the rotation of the manipulation space. This defines
the (orthonormally-constrained) rotation on the basis coefficients.

For a radial tour, fix the (horizontal) angle within the projection
plane, \(\theta\), and define a sequence for the (vertical) angle,
\(\phi\), defining movement out of the plane, that changes the initial
value of \(\phi_1\), the angle between \(\textbf{e}\) and its shadow in
\(\textbf{B}\), to a maximum of 1, and then to 0, before returning to
\(\phi_1\).

Rotations in 3D can be defined by the axes they pivot on. \(\theta\)
describes the rotation around the \(Z\) axis. While \(\phi\) is the
rotation around \(\textbf{U}\), an axis on the \(XY\) plane, orthogonal
to \(\textbf{e}\). Given these axes, the SO(3) rotation matrix can be
derived as follows:

@@

\hypertarget{creating-the-animation-of-the-radial-rotation}{%
\subsection{Creating the animation of the radial
rotation}\label{creating-the-animation-of-the-radial-rotation}}

The steps outlined above can be used to create any arbitrary rotation in
the manip space. To use these for sensitivity analysis, the radial
rotation is built into an animation where the manip var is rotated fully
into the projection, completely out, and then back to the initial value.
This involves allowing \(\phi\), to vary between 0 and 1.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Set initial value of \(\phi_1\) and \(\theta\):
  \(\phi_1 = \cos^{-1}{\sqrt{B_{k1}^2+B_{k2}^2}}\),
  \(\theta=\tan^{-1}\frac{B_{k2}}{B_{k1}}\). \(\phi_1\) is the angle
  between \(\textbf{e}\) and its shadow in \(\textbf{B}\). Check that
  \(\phi_1\) is between 0 and \(\pi/2\), and if not, transform it into
  this range, by subtracting or adding \(\pi\) XXX.
\item
  Set an angle increment (\(\Delta_\phi\)) that sets the step size to
  make, to rotate the manip var into and out of the projection. (Note:
  Using angle increment, rather than a number of steps, to control the
  movement, is consistent with the tour algorithm as implemented in the
  \pkg{tourr}).
\item
  Step towards \(0\), where the manip var has a completely in the
  projection.
\item
  Step towards \(\pi/2\) where the manip variable has no contribution.
\item
  Step back to \(\phi_1\).
\end{enumerate}

@@

### 2019-05-29 paper/spinifex_paper.Rmd
di tweaks on plane

- $\textbf{X}$, the data, an $n \times p$ numeric matrix to be projected to 2D.
- $\textbf{B} = (B_1,~ B_2)$, any 2D orthonormal projection basis, $p \times 2$ matrix, describing the projection from $p$ to two dimensions. We will call this the "original projection" because it is the starting point for the manual tour.
- $k$, is the index of the variable to manipulate, called the "manip var".
- $|\textbf{B}_k|$, the norm of the manip var, $k$, the length of this basis vector between 0 and 1 inclusive.
- $\textbf{e}$, a 1D coordinate zero basis vector of length $p$ with the $k$-th element set to one.
- $\textbf{M} = orthonormalize(\textbf{B}|\textbf{e})$, the manipulation space, a $p \times 3$ matrix where out-of-plane rotation takes place. 
- $\theta$, the angle of in-projection-plane rotation, for example, on the reference axes; $c_\theta, s_\theta$ are the cosine and sine.
- $\phi$, the angle of out-of-projection-plane rotation, into the manipulation space; $c_\phi, s_\phi$ are the cosine and sine.
- $\textbf{U}$, the axis of rotation for $\phi$.
- $\textbf{R}$, the rotation matrix, containing the rotaions of $\theta$ and $\phi$ within the manipulation space, $\textbf{M}$.
- $\textbf{RM} = \textbf{M} * \textbf{R}$, the result of the manipulation space after the application of the rotation matrix.
- $\textbf{P} = \textbf{X} * \textbf{RM}$, the data projection, the data as projected through the rotated manipulation space.

@@

### Step 3) Generate rotation

<!-- illustration of axis manip -->
Imagine moving the axis of manip var, the red line in Figure \ref{fig:step2}, like a lever. This is effectively the process of the radial manual control, that rotates the manip variable into and out of the 2D projection (Figure \ref{fig:step3}). As the variable is controlled, the manipulation space rotates, and the projection onto the horizontal projection plane correspondingly changes. This is a manual tour. Generating a sequence of values for the horizontal and vertical angles produces a path for the rotation of the manipulation space. This defines the (orthonormally-constrained) rotation on the basis coefficients.

<!-- describe manip var path -->
For a radial tour, fix the (horizontal) angle within the projection plane, $\theta$, and define a sequence for the (vertical) angle, $\phi$, defining movement out of the plane, that changes the initial value of $|\textbf{B}_k|$ to a maximum of 1, and then to 0, before returning to the initial value.

<!-- define the rotation matrix -->
Rotations are defined by the axes they pivot on, $\theta$ then is the rotation around the $Z$ axis. While $\phi$ will be the rotation around, $\textbf{U}$, a new axis on the $XY$ plane, orthogonal to $\phi$. Given these axes, the SO(3) rotation matrix can be derived by the following.

@@

## Creating the animation

<!--  phi transform and animation -->
The steps outlined above can be used to create any arbitrary rotation in the manip space. To use these for sensitivity analysis, the radial rotation is built into an animation where the manip var is rotated fully into the projection, completely out, and then back to the initial value. This involves changing $\phi$, the angle relative to its initial value. Apply the transformation $\phi_i \leftarrow \phi_i-\phi_1$ such that $\phi$ is relative to the projection plane. When $\phi$ is 0, $|\textbf{B}_k| = 1$, and when $\phi$ is $\pi/2$, the manip var has no contribution to the projection.

@@

1. Set initial value to be $\phi_1$.
2. Set an angle value ($\Delta_\phi$) that sets the step size to make in rotating the manip var into and out of the projection. (Note: Using angle value to control this is consistent with the tour algorithm, as implemented in the \pkg{tourr}).
3. Step towards $0$, where the manip var has a completely in the projection.
4. Step towards $\pi/2$ where the manip variable has no contribution.
5. Step back to $\phi_1$.

@@

### Checks that need to be made

The value of $\phi$ is relative to its initial value. The transformation $\phi_i$ - $\phi_1$ is useful to think about $\phi$ relative to the basis plane. Additionally, the value of $\phi$ may be out of phase by a factor of $\pi$. If the rotation does not behave as expected these are the first places to check.

>>>>

- $\textbf{X}$, the data, an $n \times p$ numeric matrix to be projected to 2D.
- $\textbf{B} = (B_1,~ B_2)$, any 2D orthonormal projection basis, $p \times 2$ matrix, describing the projection from $p$ to two dimensions. We will call this the "original projection" because it is the starting point for the manual tour.
- $k$, is the index of the variable to manipulate, called the "manip var".
- $|\textbf{B}_k|$, the norm of the manip var, $k$, the length of this basis vector between 0 and 1 inclusive.
- $\textbf{e}$, a 1D coordinate zero basis vector of length $p$ with the $k$-th element set to one.
- $\textbf{M} = orthonormalize(\textbf{B}|\textbf{e})$, the manipulation space, a $p \times 3$ matrix where out-of-plane rotation takes place. 
- $\theta$, the angle of in-projection-plane rotation, for example, on the reference axes; $c_\theta, s_\theta$ are the cosine and sine.
- $\phi$, the angle of out-of-projection-plane rotation, into the manipulation space; $c_\phi, s_\phi$ are the cosine and sine. The initial value for animation purposes is $\phi_1$.
- $\textbf{U}$, the axis of rotation for $\phi$.
- $\textbf{R}$, the rotation matrix, containing the rotaions of $\theta$ and $\phi$ within the manipulation space, $\textbf{M}$.
- $\textbf{RM} = \textbf{M} * \textbf{R}$, the result of the manipulation space after the application of the rotation matrix.
- $\textbf{P} = \textbf{X} * \textbf{RM}$, the data projection, the data as projected through the rotated manipulation space.

XXX Clean up the notation, not all of this is needed or useful. The shadow of $e$ in the projection may need some notation. The projected data should be $Y$ not $P$.

@@

### Step 3) Defining a 3D rotation

<!-- illustration of axis manip -->
The basis vector corresponding to the manip var (red line in Figure \ref{fig:step2}), can be operated like a lever. This is the process of the manual control, that rotates the manip variable into and out of the 2D projection (Figure \ref{fig:step3}). As the variable is controlled, the manipulation space rotates, and the projection onto the horizontal projection plane correspondingly changes. This is a manual tour. Generating a sequence of values for the horizontal and vertical angles produces a path for the rotation of the manipulation space. This defines the (orthonormally-constrained) rotation on the basis coefficients.

<!-- describe manip var path -->
For a radial tour, fix the (horizontal) angle within the projection plane, $\theta$, and define a sequence for the (vertical) angle, $\phi$, defining movement out of the plane, that changes the initial value of $\phi_1$, the angle between $\textbf{e}$ and its shadow in $\textbf{B}$, to a maximum of 1, and then to 0, before returning to $\phi_1$.

<!-- define the rotation matrix -->
Rotations in 3D can be defined by the axes they pivot on. $\theta$ describes the rotation around the $Z$ axis. While $\phi$ is the rotation around $\textbf{U}$, an axis on the $XY$ plane, orthogonal to $\textbf{e}$. Given these axes, the SO(3) rotation matrix can be derived as follows:

@@

## Creating the animation of the radial rotation

<!--  phi transform and animation -->
The steps outlined above can be used to create any arbitrary rotation in the manip space. To use these for sensitivity analysis, the radial rotation is built into an animation where the manip var is rotated fully into the projection, completely out, and then back to the initial value. This involves allowing $\phi$, to vary between 0 and 1.

<!--Apply the transformation $\phi_i \leftarrow \phi_i-\phi_1$ such that $\phi$ is relative to the projection plane. When $\phi$ is 0, $|\textbf{B}_k| = 1$, and when $\phi$ is $\pi/2$, the manip var has no contribution to the projection.-->

@@

1. Set initial value of $\phi_1$ and $\theta$: $\phi_1 = \cos^{-1}{\sqrt{B_{k1}^2+B_{k2}^2}}$, $\theta=\tan^{-1}\frac{B_{k2}}{B_{k1}}$. $\phi_1$ is the angle between $\textbf{e}$ and its shadow in $\textbf{B}$. Check that  $\phi_1$ is between 0 and $\pi/2$, and if not,  transform it into this range, by subtracting or adding $\pi$ XXX. 
2. Set an angle increment ($\Delta_\phi$) that sets the step size to make, to rotate the manip var into and out of the projection. (Note: Using angle increment, rather than a number of steps, to control the movement, is consistent with the tour algorithm as implemented in the \pkg{tourr}).
3. Step towards $0$, where the manip var has a completely in the projection.
4. Step towards $\pi/2$ where the manip variable has no contribution.
5. Step back to $\phi_1$.

@@

<!--
### Checks that need to be made

The initial value of $\phi$ is determined by the angle between $\textbf{e}$ and its shadow in $\textbf{B}$.

relative to its initial value. The transformation $\phi_i$ - $\phi_1$ is useful to think about $\phi$ relative to the basis plane. Additionally, the value of $\phi$ may be out of phase by a factor of $\pi$. If the rotation does not behave as expected these are the first places to check.
-->
